---
title: "Income, Crime, and Borders in Canton Zurich"
author: "Thilo Holstein, Hans Josef Thalathara"
date: "2026-01-26"
output:
  pdf_document:
    latex_engine: xelatex
    toc: true
    toc_depth: '3'
    number_sections: true
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    theme: default
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(readr)
library(MASS)
library(dplyr)
library(sf)
library(ggplot2)
library(tidyr)
library(scales)
library(gridExtra)

source("Scripts/Data Preparation.R")
```

# Introduction

Burglary patterns across Swiss municipalities have long been a subject of practical interest for law enforcement agencies and local communities alike. In the Canton of Basel-Landschaft, anecdotal observations from police practitioners suggest that burglaries tend to cluster in wealthier municipalities located near Switzerland's international borders. The reasoning behind this observation is intuitive: affluent households may present more attractive targets, and proximity to a border could facilitate quick escape across national jurisdictions. If true, this pattern would have direct implications for resource allocation in crime prevention.

However, anecdotal observations -- no matter how consistent -- are not the same as empirical evidence. What appears as a clear pattern in one canton may not hold elsewhere, and apparent correlations can dissolve once confounding factors are accounted for. This analysis takes those observations as a starting point and asks: do they generalise?

Specifically, we examine whether the same relationship between income, border proximity, and burglary rates exists in the Canton of Zurich -- a canton with a different economic structure, geographic configuration, and population density than Basel-Landschaft. By combining burglary statistics, income data, and geographic boundary data at the municipal level, we test these assumptions empirically using both exploratory visualisation and formal count regression models. The goal is not to confirm the Basel-Landschaft narrative, but to evaluate it critically in a new context.

## Research Question

The central question guiding this analysis is:

> *How does income level relate to burglary rates across municipalities in Canton Zurich, and does proximity to the national border act as an independent predictor of crime?*

This question has two dimensions. First, we investigate whether wealthier municipalities systematically experience more burglaries -- a relationship that would suggest property crime follows economic incentives. Second, we test whether geographic proximity to the Swiss national border adds explanatory power beyond income, which would point to spatial or cross-border dynamics as a contributing factor.

## Approach

We analyse open data from Canton Zurich to investigate these questions at the municipal level. By joining income statistics, crime records, and geographic boundary data, we examine whether income predicts burglary rates, whether border proximity adds explanatory power, and how these patterns evolve over time. The analysis proceeds in three stages: data preparation and cleaning, exploratory visualisation to identify patterns and anomalies, and formal count regression modelling to test our hypotheses statistically.

## Hypotheses

Based on the anecdotal observations described above, we formulate three testable hypotheses. These range from the individual effects of income and border proximity to their potential interaction.

**H1 – Income and Burglaries:**
Municipalities with higher median income exhibit higher burglary rates, as wealthier areas may present more attractive targets.

**H2 – National Border Proximity and Burglaries:**
Municipalities located closer to national borders may experience higher burglary rates due to reduced distances to potential escape routes across international boundaries.

**H3 -- Combined Effect:**
Income and border proximity jointly predict burglary rates better than either variable alone.

# Data Sources

Our analysis draws on five publicly available datasets from three different sources, provided in two distinct formats (CSV and GeoPackage). All data were downloaded locally and are included in the project repository.

## Burglary Data

Burglary statistics by municipality and city district for Canton Zurich (2009--2024), published by the Canton of Zurich on the Open Data portal (data.zh.ch). The dataset contains approximately 15,000 records with variables including municipality identifiers, year, offence type, number of offences, population, and burglary rate per 1,000 inhabitants. Format: CSV.

## Income Data

Two separate income datasets were used:

- **Canton level:** Median taxable income by municipality (2009--2022), published by the Statistical Office of Canton Zurich (data.zh.ch). Format: CSV.
- **City district level:** Median taxable income by city district for the city of Zurich (2009--2023), published by the City of Zurich (data.stadt-zuerich.ch). Format: CSV.

The two datasets were joined at different spatial levels to achieve complete income coverage across both municipalities and city districts.

## Geographic Data

- **Municipality boundaries:** GeoPackage of municipal boundaries for Canton Zurich (data.zh.ch), used for spatial visualisation and as join keys for the burglary and income data.
- **City district boundaries:** GeoPackage of Zurich city district boundaries (data.stadt-zuerich.ch), providing geometries for the 12 Stadtkreise.
- **Swiss national borders:** GeoPackage from swisstopo (swissBOUNDARIES3D), used to calculate each municipality's distance to the nearest national border.

# Cleaning and Formatting the Dataset

## Data Formatting

Burglary data for the canton of Zurich were restricted to records representing total burglaries (“Einbrüche insgesamt”), which was the sum across different burglary types. Observations with unknown municipality names or unknown city districts were excluded. Variables related to legal references were removed, as they were not required for the analysis.

Median income data were integrated at two seperate spatial levels. At the municipality level, median income data from Canton Zurich were filtered to exclude cantonal aggregates and the city of Zurich itself. The dataset was reduced to essential variables: municipality identifier, name, year, and income value. At the city district level, median income data for the city of Zurich were filtered to include only the standard tax rate, and converted from thousands to actual values by multiplying the 50th percentile income by 1000.

Spatial boundary data for municipalities in the canton of Zurich were obtained from an official GeoPackage. Only geometries corresponding to municipalities were retained, while non-municipality and city-level geometries were excluded. The dataset was reduced to the municipality identifier and the corresponding geometry.

Swiss national border data were sourced from swisstopo and imported from a GeoPackage file. The data were filtered to retain only the Swiss national boundary. The polygon representation of the national territory was converted into a boundary line to enable distance calculations.

Prior to spatial analysis, all municipality geometries were transformed to a common coordinate reference system matching that of the Swiss border data. For each municipality, the centroid of the polygon geometry was computed. Euclidean distances from these centroids to the Swiss national border line were then calculated. Distances were stored in meters and additionally converted into kilometers for interpretability.

City district geometries for the city of Zurich were processed separately. Because the original dataset contained curved polygon geometries, the data were converted into standard multipolygon geometries using GDAL utilities. The converted geometries were subsequently imported, restricted to district identifiers and geometries, and transformed to the same coordinate reference system as the Swiss border data. Centroids were calculated for each city district, and distances to the Swiss national border were computed analogously to the municipality-level analysis, again expressed in both meters and kilometers.

```{r echo = FALSE, include = FALSE}
## Burglary data canton ZH
ebd <- read.csv("Data/KTZH_EBD.csv")

# Remove unnecessary records
ebd <- ebd %>% 
  filter(Tatbestand == 'Einbrüche insgesamt', 
         Gemeindename != "unbekannt ZH",
         Stadtkreis_Name != "unbekannt") %>% 
  select(-c(Gesetz_Nummer, Gesetz_Abk))

## Income data on municipality level
income_kt <- read.csv("Data/KTZH_Income_median.csv")

# Remove non-municipality and city data and unused columns
income_kt <- income_kt %>% 
  filter(!BFS_NR %in% c(0, 261)) %>% 
  select(BFS_NR, GEBIET_NAME, INDIKATOR_JAHR, INDIKATOR_VALUE) %>% 
  rename(INCOME_VALUE = INDIKATOR_VALUE)

## Income data on city district level
income_st <- read.csv("Data/STZH_Income_median.csv")

income_st <- income_st %>% 
  filter(SteuerTarifLang == "Grundtarif") %>% 
  select(-c("KreisSort", 
            "KreisCd", 
            "SteuerTarifSort", 
            "SteuerTarifCd",
            "SteuerTarifLang")) %>% 
  mutate(INCOME_VALUE = SteuerEinkommen_p50 * 1000)

# Read .gpkg of municipalities
ktzh_gpkg <- st_read("Data/KTZH_Gemeindegrenzen_OGD.gpkg", 
                             layer = "UP_GEMEINDEN_SEEN_F")

# Remove non-municipality and city geoms
ktzh_geom <- ktzh_gpkg %>% 
  filter(!ktzh_gpkg$BFS %in% c(0, 261), ktzh_gpkg$ART_CODE == 1) %>% 
  select(BFS, geom)

# Read .gpkg of swiss borders (swisstopo)
swiss_borders <- st_read("Data/CH_Borders_swisstopo.gpkg", 
                         layer = "tlm_landesgebiet")

swiss_borders <- swiss_borders %>% 
  filter(swiss_borders$icc == "CH")

# Calculate distance to border
swiss_border_line <- st_boundary(swiss_borders)

# Ensure same CRS
ktzh_geom <- st_transform(ktzh_geom, st_crs(swiss_borders))

# Calculate distance to border
ktzh_geom <- ktzh_geom %>% 
  st_transform(st_crs(swiss_borders)) %>%
  mutate(
    distance_to_border = st_distance(st_centroid(.), swiss_border_line)[,1],
    distance_to_border_km = as.numeric(distance_to_border) / 1000
  ) %>% 
  select(BFS, geom, distance_to_border, distance_to_border_km)

# Convert the city district .gpkg CURVEPOLYGON TO MULTIPOLYGON using gdal_utils
gdal_utils(
  util = "vectortranslate",
  source = "Data/STZH_Stadtkreise_OGD.gpkg",
  destination = "Data/STZH_lin.gpkg",
  options = c("-nlt", "MULTIPOLYGON", 
              "-lco", "GEOMETRY_NAME=geom",
              "-overwrite")
)

# Read the converted file
stzh_gpkg <- st_read("Data/STZH_lin.gpkg")

stzh_geom <- stzh_gpkg %>% 
  select(kname, geom) %>% 
  st_transform(st_crs(swiss_borders)) %>% 
  mutate(
    distance_to_border = st_distance(st_centroid(.), swiss_border_line)[,1],
    distance_to_border_km = as.numeric(distance_to_border) / 1000
  )
```

## Investigating Missing Values

After joining the datasets, 353 of 2,735 observations (12.9%) have missing values in the income variable (`INCOME_VALUE`). These missing values arise from three distinct causes:

- **2023 (158 missing):** Municipal income statistics for Canton Zurich had not yet been published at the time of analysis. Only the 12 city districts of Zurich have income data for this year, sourced from the separate city-level dataset.
- **2024 (171 missing):** No income data are available for any municipality or city district.
- **2009--2014 (24 missing):** Four municipalities (Stammheim, Wädenswil, Elgg, and Horgen) lack income records prior to 2015, likely due to later inclusion in the cantonal statistics.

All other variables -- burglary counts, population, burglary rates, and border distance -- are complete across the entire observation period. Since the missingness is entirely explained by data availability rather than by any systematic relationship with the outcome variable, the missing-at-random assumption is satisfied.

We retain the full dataset for descriptive and spatial analyses (maps, time series, distributions), where income is not required. For all analyses involving income as a variable -- including the correlation matrix, scatterplots, and regression models -- observations with missing income values are excluded. This approach maximises the use of available data while ensuring that statistical models are estimated on complete cases only.

## Joining the Datasets

```{r include = TRUE}
# Merge data sets
data <- ebd %>% 
  left_join(income_kt, 
            by = c("Gemeinde_BFS_Nr" = "BFS_NR",
                   "Ausgangsjahr" = "INDIKATOR_JAHR")) %>% 
  left_join(income_st %>% 
              select(KreisLang, StichtagDatJahr, INCOME_VALUE) %>% 
              rename(INCOME_VALUE_ST = INCOME_VALUE),
            by = c("Stadtkreis_Name" = "KreisLang",
                   "Ausgangsjahr" = "StichtagDatJahr")) %>% 
  mutate(INCOME_VALUE = if_else(!is.na(INCOME_VALUE_ST), 
                                INCOME_VALUE_ST, 
                                INCOME_VALUE)) %>% 
  select(-INCOME_VALUE_ST)

# Join geoms of the municipalities
data <- data %>% 
  left_join(ktzh_geom, by = c("Gemeinde_BFS_Nr" = "BFS"))

# Convert data's geometry to MULTIPOLYGON first
data <- data %>%
  mutate(geom = st_cast(geom, "MULTIPOLYGON"))

# Join and replace only the EMPTY geometries
data <- data %>%
  left_join(
    stzh_geom %>%
      select(kname, 
             geom_stzh = geom, 
             distance_to_border_stzh = distance_to_border, 
             distance_to_border_km_stzh = distance_to_border_km),
    by = c("Stadtkreis_Name" = "kname")
  ) %>%
  mutate(
    geom = if_else(st_is_empty(geom) & !is.na(geom_stzh), 
                   geom_stzh, 
                   geom),
    distance_to_border = coalesce(distance_to_border_stzh, distance_to_border),
    distance_to_border_km = coalesce(distance_to_border_km_stzh, distance_to_border_km)
  ) %>%
  select(-geom_stzh, -distance_to_border_stzh, -distance_to_border_km_stzh) %>%
  st_as_sf() %>% 
  mutate(Gemeinde_BFS_Nr = if_else(Gemeindename == "Zürich", 
                                   Stadtkreis_BFS_Nr, 
                                   Gemeinde_BFS_Nr))
```

# Exploratory Data Analysis

Before testing our hypotheses through formal modelling, we explore the data visually. This section begins with a spatial overview of burglary patterns across the canton, then follows a narrative from the macro level (canton-wide trends over time) down to the micro level (individual municipality patterns), building an increasingly detailed picture of how income, burglaries, and geography interact across Canton Zurich.

## Spatial Overview: Burglary Rates by Municipality

The following maps show the total number of burglaries per municipality, normalised by average population, across the full observation period. This provides a first visual impression of how burglary risk is distributed geographically.

```{r echo = FALSE, fig.show='hold'}
# City outline for map
# Create the outer boundary of Zurich city
city_outline <- data %>%
  filter(!is.na(Stadtkreis_BFS_Nr)) %>%
  st_union()  # Dissolves all districts into one polygon

# Get bounding box of municipalities
bbox <- st_bbox(ktzh_geom)
padding <- 5000  # in map units (meters for Swiss CRS)

# Plot the crime rate across whole data set
ebd_per_gmd <- data %>% 
  group_by(Gemeinde_BFS_Nr) %>% 
  mutate(Total_EBD = sum(Straftaten_total, na.rm = TRUE), 
         Einwohner_avg = mean(Einwohner, na.rm = TRUE),
         EBD_pop_ratio = Total_EBD / Einwohner_avg) %>% 
  summarise(across(c(geom, Einwohner_avg, Total_EBD, EBD_pop_ratio), first),
  .groups = "drop") %>% 
  st_as_sf()

ggplot() +
  geom_sf(data = swiss_borders, fill = NA, color = "black", linewidth = 1) +
  geom_sf(data = ebd_per_gmd, aes(fill = EBD_pop_ratio), color = "white", linewidth = 0.3) +
  geom_sf(data = city_outline, 
          fill = NA, 
          color = "black", 
          size = 0.5) +
  scale_fill_viridis_c(name = "EBD ratio per population") +
  coord_sf(xlim = c(bbox["xmin"] - padding, bbox["xmax"] + padding), 
           ylim = c(bbox["ymin"] - padding, bbox["ymax"] + padding)) +
  theme_minimal() +
  labs(title = "Number of Burglaries per Municipality by Population")
```

The city district Kreis 1 seems to be a strong outlier in the data. To have a more balanced output of the burglary rate per municipality and population, this is how the map of the canton of Zurich looks without the outlier:

```{r echo = FALSE}
ggplot() +
  geom_sf(data = swiss_borders, fill = NA, color = "black", linewidth = 1) +
  geom_sf(data = ebd_per_gmd %>% filter(Gemeinde_BFS_Nr != 2610001), aes(fill = EBD_pop_ratio), color = "white", linewidth = 0.3) +
  geom_sf(data = city_outline, 
          fill = NA, 
          color = "black", 
          size = 0.5) +
  scale_fill_viridis_c(name = "EBD ratio per population") +
  coord_sf(xlim = c(bbox["xmin"] - padding, bbox["xmax"] + padding), 
           ylim = c(bbox["ymin"] - padding, bbox["ymax"] + padding)) +
  theme_minimal() +
  labs(title = "Number of Burglaries per Municipality by Population (without Kreis 1)")
```

## Descriptive Overview

We begin with a summary of the dataset's structure and the distributions of our key variables. This provides the necessary context for all subsequent analyses.

```{r descriptive-overview, results='asis'}
desc_data <- data %>% st_drop_geometry()

n_obs <- nrow(desc_data)
n_municipalities <- n_distinct(desc_data$Gemeinde_BFS_Nr)
year_range <- range(desc_data$Ausgangsjahr)
n_years <- n_distinct(desc_data$Ausgangsjahr)

cat(paste0(
  "The dataset contains **", format(n_obs, big.mark = "'"), " observations** across **",
  n_municipalities, " municipalities and city districts** over **",
  n_years, " years** (", year_range[1], "–", year_range[2], "). ",
  "Each observation represents one municipality in one year, recording the number of burglaries, ",
  "population size, burglary rate, median income, and distance to the national border.\n"
))
```

```{r summary-table, results='asis'}
summary_stats <- desc_data %>%
  summarise(
    across(
      c(Straftaten_total, Einwohner, Häufigkeitszahl, INCOME_VALUE, distance_to_border_km),
      list(
        mean = ~mean(.x, na.rm = TRUE),
        median = ~median(.x, na.rm = TRUE),
        sd = ~sd(.x, na.rm = TRUE),
        min = ~min(.x, na.rm = TRUE),
        max = ~max(.x, na.rm = TRUE),
        missing = ~sum(is.na(.x))
      ),
      .names = "{.col}__{.fn}"
    )
  ) %>%
  pivot_longer(everything(),
               names_to = c("variable", "stat"),
               names_sep = "__") %>%
  pivot_wider(names_from = stat, values_from = value) %>%
  mutate(variable = recode(variable,
    "Straftaten_total" = "Burglary Count",
    "Einwohner" = "Population",
    "Häufigkeitszahl" = "Burglary Rate (per 1k)",
    "INCOME_VALUE" = "Median Income (CHF)",
    "distance_to_border_km" = "Border Distance (km)"
  ))

knitr::kable(
  summary_stats %>%
    mutate(across(where(is.numeric), ~round(.x, 1))) %>%
    rename(Variable = variable, Mean = mean, Median = median,
           SD = sd, Min = min, Max = max, Missing = missing),
  row.names = FALSE,
  caption = "Summary statistics for key variables"
)
```

The distributions of our key variables reveal important characteristics of the data:

```{r dist-burglary, fig.width=10, fig.height=4}
ggplot(desc_data, aes(x = Straftaten_total)) +
  geom_histogram(bins = 50, fill = "#2c7fb8", alpha = 0.7, color = "white") +
  scale_x_continuous(labels = comma) +
  theme_minimal(base_size = 13) +
  labs(
    title = "Distribution of Burglary Counts",
    subtitle = "Per municipality-year observation",
    x = "Number of Burglaries",
    y = "Frequency"
  )
```

The burglary count distribution is heavily right-skewed: most municipality-year observations record relatively few burglaries, while a small number of large municipalities account for the extreme values. This skew is a hallmark of count data and one reason why standard linear regression is inappropriate here — a point we return to in the modelling section.

```{r dist-income-border, fig.width=10, fig.height=4}
p_income <- ggplot(desc_data %>% filter(!is.na(INCOME_VALUE)),
                   aes(x = INCOME_VALUE)) +
  geom_histogram(bins = 40, fill = "#2c7fb8", alpha = 0.7, color = "white") +
  scale_x_continuous(labels = comma_format(big.mark = "'")) +
  theme_minimal(base_size = 12) +
  labs(title = "Median Income", x = "CHF", y = "Frequency")

p_border <- ggplot(desc_data %>% filter(!is.na(distance_to_border_km)),
                   aes(x = distance_to_border_km)) +
  geom_histogram(bins = 30, fill = "#d95f02", alpha = 0.7, color = "white") +
  theme_minimal(base_size = 12) +
  labs(title = "Distance to Border", x = "km", y = "Frequency")

p_pop <- ggplot(desc_data, aes(x = Einwohner)) +
  geom_histogram(bins = 50, fill = "#4daf4a", alpha = 0.7, color = "white") +
  scale_x_log10(labels = comma) +
  theme_minimal(base_size = 12) +
  labs(title = "Population (log scale)", x = "Inhabitants", y = "Frequency")

gridExtra::grid.arrange(p_income, p_border, p_pop, ncol = 3)
```

Income is approximately normally distributed across municipalities, with most values clustered between 30,000 and 50,000 CHF. Border distance shows a right-skewed distribution — most municipalities lie within 20–30 km of the border, with few exceeding 40 km. Population varies enormously (note the log scale): from small rural communes with fewer than 2,000 residents to Zurich city districts with over 50,000.

```{r correlation-matrix, fig.width=7, fig.height=6}
cor_data <- desc_data %>%
  select(Straftaten_total, Einwohner, Häufigkeitszahl,
         INCOME_VALUE, distance_to_border_km) %>%
  filter(complete.cases(.))

cor_matrix <- cor(cor_data)

# Rename for display
colnames(cor_matrix) <- rownames(cor_matrix) <- c(
  "Burglary\nCount", "Population", "Burglary\nRate",
  "Income", "Border\nDist."
)

# Heatmap
cor_long <- as.data.frame(as.table(cor_matrix)) %>%
  rename(Var1 = Var1, Var2 = Var2, Correlation = Freq)

ggplot(cor_long, aes(x = Var1, y = Var2, fill = Correlation)) +
  geom_tile(color = "white") +
  geom_text(aes(label = round(Correlation, 2)), size = 4) +
  scale_fill_gradient2(low = "#d95f02", mid = "white", high = "#2c7fb8",
                       midpoint = 0, limits = c(-1, 1)) +
  theme_minimal(base_size = 13) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(
    title = "Correlation Matrix of Key Variables",
    x = NULL, y = NULL
  )
```

The correlation matrix provides a first hint at the relationships we will explore in detail. Raw burglary counts correlate strongly with population — unsurprisingly, since larger municipalities have more of everything. This is precisely why we use the burglary *rate* (or count models with a population offset) rather than raw counts in our analysis. The correlations between burglary rate, income, and border distance are the ones relevant to our hypotheses and will be examined in the following sections.

## Temporal Trends: Income and Burglary Rates

We begin by examining how burglary rates and median income have evolved across Canton Zurich over time. Since the two variables operate on fundamentally different scales, we normalise both to an index where the first available year equals 100. This allows direct comparison of their relative trajectories.

```{r indexed-timeline, fig.width=10, fig.height=6}
timeline_data <- data %>%
  st_drop_geometry() %>%
  group_by(Ausgangsjahr) %>%
  summarise(
    burglary_rate = mean(Häufigkeitszahl, na.rm = TRUE),
    income = mean(INCOME_VALUE, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(!is.na(income))

base_year <- min(timeline_data$Ausgangsjahr)
base_vals <- timeline_data %>% filter(Ausgangsjahr == base_year)

timeline_data <- timeline_data %>%
  mutate(
    burglary_index = burglary_rate / base_vals$burglary_rate * 100,
    income_index = income / base_vals$income * 100
  )

timeline_long <- timeline_data %>%
  select(Ausgangsjahr, burglary_index, income_index) %>%
  pivot_longer(cols = c(burglary_index, income_index),
               names_to = "variable",
               values_to = "index") %>%
  mutate(variable = recode(variable,
                           "burglary_index" = "Burglary Rate",
                           "income_index" = "Median Income"))

ggplot(timeline_long, aes(x = Ausgangsjahr, y = index,
                          color = variable, linetype = variable)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.5) +
  geom_hline(yintercept = 100, linetype = "dashed", color = "grey50", linewidth = 0.5) +
  scale_color_manual(values = c("Burglary Rate" = "#d95f02",
                                "Median Income" = "#2c7fb8")) +
  scale_x_continuous(breaks = seq(base_year, max(timeline_data$Ausgangsjahr), 1)) +
  theme_minimal(base_size = 13) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(
    title = "Burglary Rate and Median Income Over Time",
    subtitle = paste0("Canton Zurich, indexed (", base_year, " = 100)"),
    x = "Year",
    y = "Index",
    color = NULL,
    linetype = NULL
  )
```

The indexed timeline reveals a striking divergence: while median income has risen steadily over the observation period, burglary rates have declined substantially. This inverse macro-trend already raises questions about the assumed positive relationship in H1. If higher income attracted more burglaries, we would expect both lines to move in the same direction — yet they clearly do not. This motivates a closer look at whether the pattern holds at the municipal level, or whether spatial factors such as border proximity might better explain variation in burglary rates.

## Spatial Dimension: Burglary Rates by Border Proximity

The macro-level decline in burglaries may mask important spatial heterogeneity. To investigate whether geography matters, we group municipalities into three categories based on their centroid distance to the Swiss national border and track their burglary rates over time separately.

```{r border-timeline, fig.width=10, fig.height=6}
border_timeline <- data %>%
  st_drop_geometry() %>%
  filter(!is.na(distance_to_border_km), !is.na(Häufigkeitszahl)) %>%
  mutate(border_group = cut(distance_to_border_km,
                            breaks = c(0, 10, 20, Inf),
                            labels = c("< 10 km", "10–20 km", "> 20 km"))) %>%
  group_by(Ausgangsjahr, border_group) %>%
  summarise(
    burglary_rate = mean(Häufigkeitszahl, na.rm = TRUE),
    n_municipalities = n_distinct(Gemeinde_BFS_Nr),
    .groups = "drop"
  )

ggplot(border_timeline, aes(x = Ausgangsjahr, y = burglary_rate,
                            color = border_group)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.5) +
  scale_color_manual(values = c("< 10 km" = "#e41a1c",
                                "10–20 km" = "#ff7f00",
                                "> 20 km" = "#4daf4a")) +
  scale_x_continuous(breaks = seq(base_year, max(border_timeline$Ausgangsjahr), 1)) +
  theme_minimal(base_size = 13) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(
    title = "Burglary Rate Over Time by Distance to National Border",
    subtitle = "Canton Zurich, municipalities grouped by centroid distance to border",
    x = "Year",
    y = "Burglary Rate (per 1'000 inhabitants)",
    color = "Distance to Border"
  )
```

This chart reveals a clear spatial gradient: municipalities within 10 km of the national border consistently exhibit higher burglary rates than those further inland. While all three groups follow the general downward trend observed in Section 4.1, the gap between border-proximate and interior municipalities persists across the entire observation period. This suggests that border proximity is a structurally relevant factor — not merely a transient pattern — and provides initial support for H2.

## Cross-Sectional Analysis: Income vs. Burglary Rate

Having established the temporal and spatial context, we now examine the core relationship hypothesised in H1: does higher municipal income predict higher burglary rates? We aggregate each municipality across all available years and fit a simple linear regression.

```{r h1-scatterplot, fig.width=10, fig.height=7}
h1_data <- data %>%
  st_drop_geometry() %>%
  group_by(Gemeinde_BFS_Nr, Gemeindename) %>%
  summarise(
    income_median = median(INCOME_VALUE, na.rm = TRUE),
    haeufigkeit_mean = mean(Häufigkeitszahl, na.rm = TRUE),
    einwohner_mean = mean(Einwohner, na.rm = TRUE),
    distance_to_border_km = first(distance_to_border_km),
    n_years = n(),
    .groups = "drop"
  ) %>%
  filter(!is.na(income_median), !is.na(haeufigkeit_mean))

lm_h1 <- lm(haeufigkeit_mean ~ income_median, data = h1_data)

r2 <- summary(lm_h1)$r.squared
p_val <- summary(lm_h1)$coefficients[2, 4]
label_text <- paste0("R² = ", round(r2, 3),
                     "\np = ", formatC(p_val, format = "e", digits = 2))

ggplot(h1_data, aes(x = income_median, y = haeufigkeit_mean)) +
  geom_point(aes(size = einwohner_mean), alpha = 0.5, color = "#2c7fb8") +
  geom_smooth(method = "lm", se = TRUE, color = "#d95f02", linewidth = 1) +
  scale_x_continuous(labels = comma_format(big.mark = "'")) +
  scale_size_continuous(name = "Population (avg.)", labels = comma) +
  annotate("text", x = Inf, y = Inf, label = label_text,
           hjust = 1.1, vjust = 1.5, size = 4, fontface = "italic") +
  theme_minimal(base_size = 13) +
  labs(
    title = "H1: Median Income vs. Burglary Rate",
    subtitle = "Per municipality/city district, aggregated across all years",
    x = "Median Income (CHF)",
    y = "Burglary Rate (per 1'000 inhabitants)"
  )
```

The scatterplot does not support H1. The regression line slopes slightly *downward*, suggesting that wealthier municipalities tend to have marginally lower — not higher — burglary rates. However, the effect is weak: income explains only about 2% of the variance in burglary rates, and the p-value does not reach conventional significance at the 5% level. A few outliers with very high burglary rates (likely small municipalities where a handful of incidents inflate the rate) are visible in the upper portion of the plot. Given these findings, income alone is a poor predictor of burglary rates. This motivates the inclusion of border proximity as an additional predictor.

## Municipality Typology: Income--Crime Clusters

We classify all municipalities into four quadrants based on whether their median income and average burglary rate fall above or below the canton-wide median. This typology provides an intuitive summary of the landscape and highlights which municipalities defy the expected patterns.

```{r cluster-plot, fig.width=10, fig.height=7}
cluster_data <- data %>%
  st_drop_geometry() %>%
  group_by(Gemeinde_BFS_Nr, Gemeindename, Stadtkreis_Name) %>%
  summarise(
    income_median = median(INCOME_VALUE, na.rm = TRUE),
    haeufigkeit_mean = mean(Häufigkeitszahl, na.rm = TRUE),
    einwohner_mean = mean(Einwohner, na.rm = TRUE),
    distance_to_border_km = first(distance_to_border_km),
    .groups = "drop"
  ) %>%
  filter(!is.na(income_median), !is.na(haeufigkeit_mean)) %>%
  mutate(display_name = if_else(Gemeindename == "Zürich",
                                paste0("Zürich - ", Stadtkreis_Name),
                                Gemeindename))

income_threshold <- median(cluster_data$income_median)
crime_threshold <- median(cluster_data$haeufigkeit_mean)

cluster_data <- cluster_data %>%
  mutate(cluster = case_when(
    income_median >= income_threshold & haeufigkeit_mean >= crime_threshold
      ~ "High Income / High Crime",
    income_median >= income_threshold & haeufigkeit_mean < crime_threshold
      ~ "High Income / Low Crime",
    income_median < income_threshold & haeufigkeit_mean >= crime_threshold
      ~ "Low Income / High Crime",
    income_median < income_threshold & haeufigkeit_mean < crime_threshold
      ~ "Low Income / Low Crime"
  ))

ggplot(cluster_data, aes(x = income_median, y = haeufigkeit_mean,
                         color = cluster)) +
  geom_point(aes(size = einwohner_mean), alpha = 0.6) +
  geom_vline(xintercept = income_threshold, linetype = "dashed", color = "grey40") +
  geom_hline(yintercept = crime_threshold, linetype = "dashed", color = "grey40") +
  geom_text(data = cluster_data %>%
              filter(haeufigkeit_mean > quantile(haeufigkeit_mean, 0.95) |
                     einwohner_mean > quantile(einwohner_mean, 0.95)),
            aes(label = display_name),
            size = 3, nudge_y = 0.5, show.legend = FALSE) +
  scale_color_manual(values = c(
    "High Income / High Crime" = "#e41a1c",
    "High Income / Low Crime"  = "#4daf4a",
    "Low Income / High Crime"  = "#ff7f00",
    "Low Income / Low Crime"   = "#377eb8"
  )) +
  scale_size_continuous(name = "Population (avg.)", labels = comma) +
  scale_x_continuous(labels = comma_format(big.mark = "'")) +
  theme_minimal(base_size = 13) +
  labs(
    title = "Municipality Clusters: Income vs. Burglary Rate",
    subtitle = "Dashed lines = median thresholds across all municipalities",
    x = "Median Income (CHF)",
    y = "Burglary Rate (per 1'000 inhabitants)",
    color = "Cluster"
  )
```

The quadrant plot reveals that municipalities are distributed relatively evenly across all four clusters, with no strong concentration in the "High Income / High Crime" quadrant that H1 would predict. The labelled outliers — municipalities with exceptionally high burglary rates or large populations — provide concrete cases for further investigation. Notably, the distribution of municipalities across clusters supports the earlier regression findings: income alone does not systematically sort municipalities into high- or low-crime categories. Other factors, particularly geographic location, appear to play a more decisive role.

# Modelling

The exploratory analysis revealed visual patterns -- a spatial gradient near the border and no clear income--crime link -- but visual impressions can be misleading. To rigorously test our three hypotheses, we now turn to formal statistical modelling. We move beyond simple linear regression (which treats burglary *rates* as continuous) and instead model the raw **burglary counts** directly. This is more appropriate because:

- Burglary counts are non-negative integers — classic **count data**.
- Municipalities differ vastly in population size, so we need to account for this.
- The variance in crime counts often exceeds the mean (**overdispersion**), which standard linear models cannot handle.

We fit two types of count regression models — **Poisson** and **Negative Binomial** — and compare their performance.

## Why Count Regression?

A standard linear regression on rates (Häufigkeitszahl) treats every municipality equally, regardless of whether it has 2,000 or 50,000 residents. A municipality with 2 burglaries and 1,000 residents gets the same weight as one with 200 burglaries and 100,000 residents — even though the latter estimate is far more precise. Count regression models solve this by working with the raw counts and using a so-called **offset variable** to account for population size.

### The Offset Variable

The key idea: we model `Straftaten_total` (raw burglary count) as the outcome, but include `log(Einwohner)` as an **offset**. Mathematically:

$$\log(\text{count}) = \beta_0 + \beta_1 \cdot \text{income} + \beta_2 \cdot \text{border distance} + \log(\text{population})$$

Rearranging:

$$\log\left(\frac{\text{count}}{\text{population}}\right) = \beta_0 + \beta_1 \cdot \text{income} + \beta_2 \cdot \text{border distance}$$

So the model effectively predicts the **rate** (burglaries per capita), but it correctly weights each observation by its population size. The offset is not estimated — it enters the model with a fixed coefficient of 1.

### Covariates

We include three covariates (predictor variables):

- **Income** (in 10,000 CHF): Tests whether wealthier municipalities experience more burglaries (H1).
- **Distance to border** (km): Tests whether proximity to the Swiss national border increases burglaries (H2).
- **Year** (centered): Controls for the canton-wide declining trend in burglaries over time.

```{r model-data-prep}
model_data <- data %>%
  st_drop_geometry() %>%
  filter(
    !is.na(Straftaten_total),
    !is.na(INCOME_VALUE),
    !is.na(distance_to_border_km),
    Einwohner > 0
  ) %>%
  mutate(
    log_pop = log(Einwohner),
    income_10k = INCOME_VALUE / 10000,
    year_centered = Ausgangsjahr - min(Ausgangsjahr)
  )
```

## Poisson Regression

The Poisson model is the natural starting point for count data. It assumes that the outcome follows a Poisson distribution, where the mean equals the variance (**equidispersion**).

```{r poisson-models}
poisson_m1 <- glm(
  Straftaten_total ~ income_10k + offset(log_pop),
  family = poisson(link = "log"),
  data = model_data
)

poisson_m2 <- glm(
  Straftaten_total ~ distance_to_border_km + offset(log_pop),
  family = poisson(link = "log"),
  data = model_data
)

poisson_full <- glm(
  Straftaten_total ~ income_10k + distance_to_border_km + year_centered +
    offset(log_pop),
  family = poisson(link = "log"),
  data = model_data
)
```

```{r poisson-results, results='asis'}
s_pois <- summary(poisson_full)
pois_coefs <- s_pois$coefficients

cat("**Poisson Full Model: Burglary Count ~ Income + Border Distance + Year + offset(log Population)**\n\n")

pois_table <- data.frame(
  Variable = c("(Intercept)", "Income (10k CHF)", "Distance to Border (km)", "Year"),
  Estimate = round(pois_coefs[, "Estimate"], 4),
  `Std. Error` = round(pois_coefs[, "Std. Error"], 4),
  `p-value` = formatC(pois_coefs[, "Pr(>|z|)"], format = "e", digits = 2),
  check.names = FALSE
)
knitr::kable(pois_table, row.names = FALSE)
```

Under the Poisson model, all three predictors appear highly significant. However, before interpreting these results, we must verify the model's core assumption — that the variance of burglary counts equals the mean.

### Checking for Overdispersion

The critical Poisson assumption is that the variance equals the mean. We check this with the **dispersion statistic**: the sum of squared Pearson residuals divided by the degrees of freedom. A value near 1 indicates the assumption holds; values much greater than 1 signal **overdispersion**.

```{r overdispersion-check, results='asis'}
pearson_resid <- residuals(poisson_full, type = "pearson")
dispersion <- sum(pearson_resid^2) / poisson_full$df.residual

cat(paste0("**Dispersion statistic: ", round(dispersion, 2), "**\n\n"))

if (dispersion > 2) {
  cat("The dispersion statistic is substantially greater than 1, confirming that the data are **overdispersed**. ")
  cat("This means the Poisson model underestimates the true variability in burglary counts. ")
  cat("Consequently, the standard errors are too small, and the p-values are artificially low — ")
  cat("the model appears more confident in its estimates than it should be. ")
  cat("This motivates fitting a Negative Binomial model, which explicitly accounts for overdispersion.\n")
} else {
  cat("The dispersion statistic is close to 1, suggesting that the Poisson assumption is reasonable.\n")
}
```

## Negative Binomial Regression

The Negative Binomial (NB) model extends the Poisson by adding an extra parameter **theta** ($\theta$) that captures the overdispersion. When $\theta \to \infty$, the NB reduces to the Poisson. When $\theta$ is small, there is substantial extra variability beyond what the Poisson allows.

This extra variability reflects real-world factors not captured by our covariates: differences in local policing, urban vs. rural character, socioeconomic conditions, and other unobserved heterogeneity between municipalities.

```{r nb-models}
nb_m1 <- glm.nb(
  Straftaten_total ~ income_10k + offset(log_pop),
  data = model_data
)

nb_m2 <- glm.nb(
  Straftaten_total ~ distance_to_border_km + offset(log_pop),
  data = model_data
)

nb_m3 <- glm.nb(
  Straftaten_total ~ income_10k + distance_to_border_km + offset(log_pop),
  data = model_data
)

nb_full <- glm.nb(
  Straftaten_total ~ income_10k + distance_to_border_km + year_centered +
    offset(log_pop),
  data = model_data
)

nb_interaction <- glm.nb(
  Straftaten_total ~ income_10k * distance_to_border_km + year_centered +
    offset(log_pop),
  data = model_data
)
```

### NB Model Results

```{r nb-results, results='asis'}
s_nb <- summary(nb_full)
nb_coefs <- s_nb$coefficients

cat("**NB Full Model: Burglary Count ~ Income + Border Distance + Year + offset(log Population)**\n\n")

nb_table <- data.frame(
  Variable = c("(Intercept)", "Income (10k CHF)", "Distance to Border (km)", "Year"),
  Estimate = round(nb_coefs[, "Estimate"], 4),
  `Std. Error` = round(nb_coefs[, "Std. Error"], 4),
  `p-value` = formatC(nb_coefs[, "Pr(>|z|)"], format = "e", digits = 2),
  check.names = FALSE
)
knitr::kable(nb_table, row.names = FALSE)

cat(paste0("\n\n**Theta (overdispersion parameter): ", round(nb_full$theta, 3), "**\n\n"))
```

Compared to the Poisson model, the NB standard errors are noticeably larger and several p-values increase substantially. This is the expected consequence of accounting for overdispersion: the model no longer overstates its confidence. The theta parameter quantifies the degree of extra-Poisson variability — a small theta indicates substantial overdispersion.

### Incidence Rate Ratios

The exponentiated coefficients give **Incidence Rate Ratios (IRR)**. An IRR of 0.95 for border distance means: each additional kilometre away from the border *multiplies* the expected burglary rate by 0.95 — i.e., a 5% decrease per kilometre.

```{r irr-table, results='asis'}
irr <- exp(coef(nb_full))
irr_ci <- exp(confint(nb_full))
irr_table <- data.frame(
  Variable = c("(Intercept)", "Income (10k CHF)", "Distance to Border (km)", "Year"),
  IRR = round(irr, 4),
  `CI lower (2.5%)` = round(irr_ci[, 1], 4),
  `CI upper (97.5%)` = round(irr_ci[, 2], 4),
  check.names = FALSE
)
knitr::kable(irr_table, row.names = FALSE)
```

The IRR table translates the model coefficients into a more intuitive scale. An IRR close to 1 indicates no meaningful effect. The confidence intervals for income and border distance both span 1, reinforcing that neither variable is a reliable predictor. The year variable, by contrast, has an IRR consistently below 1, confirming the steady annual decline in burglary rates.

### Interaction Model: Does Income Matter More Near the Border?

To test H3, we add an interaction term between income and border distance. If the interaction is significant, it would mean that the effect of income on burglaries *depends* on how close a municipality is to the border.

```{r interaction-results, results='asis'}
s_int <- summary(nb_interaction)
int_coefs <- s_int$coefficients

int_table <- data.frame(
  Variable = c("(Intercept)", "Income (10k CHF)", "Distance to Border (km)",
               "Year", "Income x Border Distance"),
  Estimate = round(int_coefs[, "Estimate"], 4),
  `Std. Error` = round(int_coefs[, "Std. Error"], 4),
  `p-value` = formatC(int_coefs[, "Pr(>|z|)"], format = "e", digits = 2),
  check.names = FALSE
)
knitr::kable(int_table, row.names = FALSE)
```

The interaction term (Income x Border Distance) is not significant, indicating that the relationship between income and burglary rates does not depend on how close a municipality is to the border. This means that income and border proximity operate independently — or, more precisely, that neither variable has a significant effect regardless of the other's value.

## Model Comparison: Poisson vs. Negative Binomial

Having fitted both model families, we now systematically compare them to determine which provides the more reliable basis for inference. This comparison proceeds through four complementary lenses: a formal statistical test, an information criterion, a visual comparison of coefficient uncertainty, and diagnostic checks of the residuals.

### Likelihood Ratio Test

Since the Poisson model is nested within the NB (the Poisson is a special case when $\theta \to \infty$), we can use a **likelihood ratio test** to determine whether the extra parameter $\theta$ significantly improves the fit.

```{r lr-test, results='asis'}
lr_stat <- 2 * (logLik(nb_full) - logLik(poisson_full))
lr_pval <- pchisq(as.numeric(lr_stat), df = 1, lower.tail = FALSE)

cat(paste0("- **LR statistic:** ", round(as.numeric(lr_stat), 2), "\n"))
cat(paste0("- **p-value:** ", formatC(lr_pval, format = "e", digits = 3), "\n\n"))

if (lr_pval < 0.05) {
  cat("The test is highly significant: the NB model fits substantially better than the Poisson. ")
  cat("This confirms that overdispersion is present and must be accounted for.\n")
} else {
  cat("The test is not significant: the Poisson model is adequate.\n")
}
```

### AIC Comparison

```{r aic-comparison, results='asis'}
aic_table <- data.frame(
  Model = c("Poisson (full)", "Negative Binomial (full)", "NB with interaction"),
  AIC = round(c(AIC(poisson_full), AIC(nb_full), AIC(nb_interaction)), 1),
  `Log-Likelihood` = round(c(as.numeric(logLik(poisson_full)),
                              as.numeric(logLik(nb_full)),
                              as.numeric(logLik(nb_interaction))), 1),
  check.names = FALSE
)
knitr::kable(aic_table, row.names = FALSE)

cat("\n*Lower AIC indicates better fit, penalised for model complexity.*\n")
```

### Coefficient Comparison

A key consequence of overdispersion: the Poisson model produces **artificially small standard errors**, making predictors appear more significant than they truly are. The NB model provides more honest uncertainty estimates.

```{r coef-comparison, results='asis', fig.width=10, fig.height=5}
vars <- c("income_10k", "distance_to_border_km", "year_centered")
var_labels <- c("Income (10k CHF)", "Distance to Border (km)", "Year")

comp_data <- data.frame(
  Variable = rep(var_labels, 2),
  Model = rep(c("Poisson", "Negative Binomial"), each = 3),
  Estimate = c(pois_coefs[vars, "Estimate"], nb_coefs[vars, "Estimate"]),
  SE = c(pois_coefs[vars, "Std. Error"], nb_coefs[vars, "Std. Error"])
)

ggplot(comp_data, aes(x = Variable, y = Estimate, color = Model)) +
  geom_point(position = position_dodge(width = 0.5), size = 3) +
  geom_errorbar(aes(ymin = Estimate - 1.96 * SE, ymax = Estimate + 1.96 * SE),
                position = position_dodge(width = 0.5), width = 0.2) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  scale_color_manual(values = c("Poisson" = "#d95f02", "Negative Binomial" = "#2c7fb8")) +
  coord_flip() +
  theme_minimal(base_size = 13) +
  labs(
    title = "Coefficient Estimates: Poisson vs. Negative Binomial",
    subtitle = "Error bars = 95% confidence intervals",
    x = NULL,
    y = "Coefficient Estimate",
    color = NULL
  )
```

The coefficient estimates are similar between both models, but the NB confidence intervals are substantially wider — reflecting the true uncertainty in the data. Any predictor that is "significant" under both models can be considered robustly supported.

### Observed vs. Predicted

```{r obs-vs-pred, fig.width=10, fig.height=5}
model_data <- model_data %>%
  mutate(
    pred_poisson = predict(poisson_full, type = "response"),
    pred_nb = predict(nb_full, type = "response")
  )

pred_long <- model_data %>%
  select(Straftaten_total, pred_poisson, pred_nb) %>%
  pivot_longer(cols = c(pred_poisson, pred_nb),
               names_to = "model",
               values_to = "predicted") %>%
  mutate(model = recode(model,
                        "pred_poisson" = "Poisson",
                        "pred_nb" = "Negative Binomial"))

ggplot(pred_long, aes(x = predicted, y = Straftaten_total, color = model)) +
  geom_point(alpha = 0.3, size = 1.5) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey40") +
  scale_x_log10() +
  scale_y_log10() +
  facet_wrap(~ model) +
  scale_color_manual(values = c("Poisson" = "#d95f02", "Negative Binomial" = "#2c7fb8")) +
  theme_minimal(base_size = 13) +
  labs(
    title = "Observed vs. Predicted Burglary Counts",
    subtitle = "Dashed line = perfect prediction (log-log scale)",
    x = "Predicted Count",
    y = "Observed Count"
  ) +
  theme(legend.position = "none")
```

If the models predicted perfectly, all points would fall on the diagonal dashed line. Both models show a similar pattern: predictions cluster along the diagonal for moderate counts, but diverge for very small and very large values. This is expected — municipalities with extreme burglary counts (either very few or very many) are inherently harder to predict with only three covariates. The key observation is that both models produce comparable predictions; the advantage of the NB lies not in better point predictions, but in more honest uncertainty estimates.

### Residual Diagnostics

```{r residual-diagnostics, fig.width=10, fig.height=5}
model_data <- model_data %>%
  mutate(
    resid_poisson = residuals(poisson_full, type = "pearson"),
    resid_nb = residuals(nb_full, type = "pearson")
  )

resid_long <- model_data %>%
  select(resid_poisson, resid_nb) %>%
  pivot_longer(everything(),
               names_to = "model",
               values_to = "residual") %>%
  mutate(model = recode(model,
                        "resid_poisson" = "Poisson",
                        "resid_nb" = "Negative Binomial"))

ggplot(resid_long, aes(x = residual)) +
  geom_histogram(aes(y = after_stat(density)), bins = 50,
                 fill = "#2c7fb8", alpha = 0.7) +
  stat_function(fun = dnorm, color = "#d95f02", linewidth = 1) +
  facet_wrap(~ model, scales = "free_x") +
  coord_cartesian(xlim = c(-5, 15)) +
  theme_minimal(base_size = 13) +
  labs(
    title = "Distribution of Pearson Residuals",
    subtitle = "Orange curve = standard normal (expected for a well-fitting model)",
    x = "Pearson Residual",
    y = "Density"
  )
```

```{r residual-summary, results='asis'}
cat("| Model | Mean | SD |\n")
cat("|-------|------|----|\n")
cat(paste0("| Poisson | ", round(mean(model_data$resid_poisson), 3),
           " | ", round(sd(model_data$resid_poisson), 3), " |\n"))
cat(paste0("| Negative Binomial | ", round(mean(model_data$resid_nb), 3),
           " | ", round(sd(model_data$resid_nb), 3), " |\n"))
cat("\n*For a well-fitting model: mean ≈ 0, SD ≈ 1.*\n")
```

The residual distributions confirm the earlier findings. The Poisson residuals have a standard deviation far exceeding 1, indicating systematic underfitting of the variance — a direct consequence of the equidispersion assumption being violated. The NB residuals are much closer to the expected standard normal shape, with a standard deviation nearer to 1, indicating that the extra theta parameter successfully absorbs the excess variability.

Across all comparison criteria -- the likelihood ratio test, AIC, coefficient uncertainty, and residual diagnostics -- the Negative Binomial model consistently outperforms the Poisson. The Poisson model's assumption of equidispersion is clearly violated, leading to artificially narrow confidence intervals and overly optimistic p-values. We therefore adopt the **Negative Binomial full model** as our primary model for all subsequent hypothesis tests.

## Hypothesis Assessment

With the Negative Binomial model established as the appropriate framework, we now formally evaluate each of our three hypotheses based on its coefficient estimates and significance levels.

```{r hypothesis-assessment, results='asis'}
nb_p_income <- nb_coefs["income_10k", "Pr(>|z|)"]
nb_p_border <- nb_coefs["distance_to_border_km", "Pr(>|z|)"]
nb_p_year <- nb_coefs["year_centered", "Pr(>|z|)"]
int_p <- int_coefs["income_10k:distance_to_border_km", "Pr(>|z|)"]

cat("**H1 — Income and Burglaries:** ")
if (nb_p_income < 0.05) {
  direction <- ifelse(nb_coefs["income_10k", "Estimate"] > 0, "positive", "negative")
  cat(paste0("Significant (p = ", formatC(nb_p_income, format = "e", digits = 2),
             "). The relationship is **", direction, "** — ",
             ifelse(direction == "positive",
                    "wealthier municipalities do show higher burglary rates, supporting H1.",
                    "wealthier municipalities actually show *lower* burglary rates, contradicting H1."),
             "\n\n"))
} else {
  cat(paste0("Not significant (p = ", formatC(nb_p_income, format = "e", digits = 2),
             ") in the NB model. Income is not a reliable predictor of burglary rates ",
             "when overdispersion is properly accounted for. **H1 is not supported.**\n\n"))
}

cat("**H2 — Border Proximity and Burglaries:** ")
if (nb_p_border < 0.05) {
  cat(paste0("Significant (p = ", formatC(nb_p_border, format = "e", digits = 2),
             "). Each kilometre further from the border is associated with a change in ",
             "burglary rate by a factor of ", round(exp(nb_coefs["distance_to_border_km", "Estimate"]), 4),
             ". **H2 is supported.**\n\n"))
} else {
  cat(paste0("Not significant (p = ", formatC(nb_p_border, format = "e", digits = 2),
             ") in the NB model. **H2 is not supported.**\n\n"))
}

cat("**H3 — Combined Effect:** ")
if (int_p < 0.05) {
  cat(paste0("The interaction between income and border distance is significant (p = ",
             formatC(int_p, format = "e", digits = 2),
             "). The effect of income on burglaries depends on border proximity. **H3 is supported.**\n"))
} else {
  cat(paste0("The interaction between income and border distance is not significant (p = ",
             formatC(int_p, format = "e", digits = 2),
             "). While both variables may contribute independently, their combined effect ",
             "does not improve the model beyond their individual contributions. ",
             "**H3 is not supported in its interaction form.**\n"))
}
```

# Chapter of Choice

# Conclusions

## Key Findings

This analysis set out to test whether anecdotal observations from Basel-Landschaft -- that burglaries concentrate in wealthier, border-proximate municipalities -- hold in the Canton of Zurich. The results paint a more nuanced picture than expected.

**Income is not a meaningful predictor of burglary rates (H1 not supported).** Contrary to the intuition behind H1, wealthier municipalities in Canton Zurich do not experience systematically higher burglary rates. The exploratory scatterplot showed no meaningful relationship, and the Negative Binomial regression confirmed that income is not statistically significant once overdispersion is properly accounted for. This suggests that the "wealthy municipalities attract burglars" narrative, while intuitively appealing, does not hold up to empirical scrutiny in this context.

**Border proximity is not a significant predictor in the formal model (H2 not supported).** While the exploratory analysis revealed a visually striking spatial gradient -- municipalities within 10 km of the border consistently showing higher burglary rates than those further inland -- this pattern does not survive formal modelling. In the Negative Binomial regression, border distance is not statistically significant (p = 0.27). This discrepancy between the visual pattern and the model result likely reflects the high variability between municipalities: while the *average* trend points in the expected direction, the municipality-to-municipality variation is too large for the effect to be statistically reliable. The border proximity pattern observed in the EDA may be confounded with other spatial factors such as urbanisation or transport connectivity.

**The combined effect of income and border proximity does not improve prediction (H3 not supported).** The interaction model did not yield a significant interaction term. Neither income nor border proximity -- individually or in combination -- are reliable predictors of burglary rates in the Negative Binomial framework.

**Burglary rates have declined substantially over time.** Across all municipalities and border distance categories, burglary rates have fallen markedly since 2009. The year variable is the strongest and most robust predictor in the model, capturing a canton-wide downward trend that likely reflects broader factors such as improved security measures, changes in policing strategies, or shifts in criminal behaviour. This temporal decline is the clearest finding of the analysis.

## Limitations

Several limitations should be considered when interpreting these results.

**Ecological fallacy.** All analyses operate at the municipal level. Even where visual patterns suggest spatial gradients (e.g., higher burglary rates near the border), we cannot draw conclusions about individual-level behaviour from aggregate data.

**Centroid distance as a proxy.** Border distance is measured from each municipality's geographic centroid to the nearest national border. This is a simplification -- municipalities with irregular shapes or large areas may have actual border distances that differ substantially from the centroid measure. A more granular measure might yield different results.

**Omitted variables.** Neither income nor border distance proved to be significant predictors in the formal model, yet the exploratory analysis showed clear visual patterns. This suggests that important explanatory variables -- such as urbanisation, transport connectivity, policing intensity, or local socioeconomic structure -- are missing from the model and may confound the observed spatial patterns.

**Incomplete income data.** Income data are unavailable for 2023--2024 and for four municipalities prior to 2015, limiting the temporal scope of income-related analyses to 2009--2022. While this affects only the income-dependent analyses and is documented in Section 3.2, it means our models cannot capture the most recent developments.

**Single canton.** The analysis is restricted to Canton Zurich. The patterns observed here may not generalise to other cantons, particularly those with different border configurations (e.g., Basel-Landschaft, Geneva, or Ticino).

# Generative AI Reflection
